A Comparsion of Traffic Engineering Methods for Application Perceived Metrics

### 2015-07-17 shrink/background.tex
added abstract, small reference fixes in rest of thesis.
Some edits to shrink chapter

\section{Content datacenter background}
\label{sec:background}

@@

\begin{figure}[t]
\centering
\begin{small}
\begin{tabular}{| m{0.9cm} | m{0.7cm} | m{3cm} | m{1.8cm} |}
\hline
Topo-logy & Server count & Switches & Switch pow. / Server pow. \\ \hline
\multicolumn{4}{|c|}{1 Gbps server link}                                                  \\\hline
FatTree  & 3456        & Cisco 2224TP (80W, 720 count) & 0.17                                      \\\hline
VL2        & 2880        & Cisco  2224TP (80W, 144 count), Cisco Nexus 5548P (390 W, 24 count)                               & 0.08                                      \\ \hline
\multicolumn{4}{|c|}{10 Gbps server link}                                                  \\\hline
FatTree  & 3456        & Cisco Nexus 5548P (390 W, 360 count) & 0.40                                      \\\hline
VL2        & 2304        & Cisco 6001 (750W, 48 count), Cisco 6004 (2900W, 6 count) & 0.23                                      \\ \hline
\end{tabular}
\end{small}
\caption{Ratio of network to server energy use at typical operating conditions. Switch power use data from Cisco \cite{cisco-dc-switches}. Server power use  of Acer Altos T350 F2 at a load of 30\% is 98.5 W \cite{spec}.}
\label{fig:network-power-use}
\end{figure}

%\textbf{Why datacenters, including \cdc s,  should reduce their network energy use?}
Datacenters, including \cdc s, should reduce their network energy use as it consumes a non-trivial fraction of datacenter energy. As shown in Figure \ref{fig:network-power-use},  switches in the FatTree \cite{fattree} and VL2 \cite{vl2} toplogies, which support full bisection bandwidth, consume 8\% and 17\% of the server energy use  at typical operating conditions in networks with a 1 Gbps server link capacity. In networks with a 10 Gbps server link capacity, which are gaining adoption, this fraction increases to 40\% and 23\% for FatTree and VL2 respectively.

>>>>

\section{Content datacenter background}
\label{sec:shrink-background}

@@

\begin{figure}[t]
\centering
\small
\begin{tabular}{| m{1.5cm} | m{1.5cm} | m{7cm} | m{3cm} |}
\hline
Topology & Server count & Switches & Switch pow. / Server pow. \\ \hline
\multicolumn{4}{|c|}{1 Gbps server link}                                                  \\\hline
FatTree  & 3456        & Cisco 2224TP (80W, 720 count) & 0.17                                      \\\hline
VL2        & 2880        & Cisco  2224TP (80W, 144 count), Cisco Nexus 5548P (390 W, 24 count)                               & 0.08                                      \\ \hline
\multicolumn{4}{|c|}{10 Gbps server link}                                                  \\\hline
FatTree  & 3456        & Cisco Nexus 5548P (390 W, 360 count) & 0.40                                      \\\hline
VL2        & 2304        & Cisco 6001 (750W, 48 count), Cisco 6004 (2900W, 6 count) & 0.23                                      \\ \hline
\end{tabular}

\caption{Ratio of network to server energy use at typical operating conditions. Switch power use data from Cisco \cite{cisco-dc-switches}. Server power use  of Acer Altos T350 F2 at a load of 30\% is 98.5 W \cite{spec}.}
\label{fig:network-power-use}
\end{figure}

%\textbf{Why datacenters, including \cdc s,  should reduce their network energy use?}
Datacenters, including \cdc s, should reduce their network energy use as it consumes a non-trivial fraction of datacenter energy. As shown in Figure \ref{fig:network-power-use},  switches in the FatTree \cite{fattree} and VL2 \cite{vl2} toplogies, which support full bisection bandwidth, consume 17\% and 8\% of the server energy use  at typical operating conditions in networks with a 1 Gbps server link capacity. In networks with a 10 Gbps server link capacity, which are gaining adoption, this fraction increases to 40\% and 23\% for FatTree and VL2 respectively.

### 2015-07-17 shrink/conclusions.tex
Some edits to shrink chapter

%!TEX root = shrink.tex
\section{Conclusions}

Despite much prior work on reducing the energy use of datacenters via server and network consolidation, a major barrier to their widespread adoption today is that operators worry deeply about the impact on SLAs or user-perceived response times, but the precise relationship between energy savings and response times is not well understood. Our work takes a step towards addressing this concern by presenting a model to quantify the energy savings vs. response time inflation trade-off in \cdc s. Based on these insights, we have designed and implemented \shrink, a system that leverages this tradeoff to yield significant energy savings while affecting user-perceived response times in a controlled manner. Shrink's novel network-aware server consolidation algorithm reduces network energy use by up to 42\% compared to network-unaware server consolidation schemes. Our experiments based on content access traces from an Akamai datacenter showed that our energy optimization algorithms reduce energy use by 35\% compared to a conservative baseline scheme that provisions servers as per the peak demand while increasing mean response time by 8\% over it. Overall, our findings encourage deployment of energy optimization techniques  in \cdc s.
%Despite much work on reducing the energy use of datacenters via server and network consolidation, a lack of quantitative understanding of energy vs. response time tradeoff is potentially a major roadblock to widespread deployment of energy saving schemes in \cdc s. Our work addressed this concern by quantifying the tradeoff between end-user response time impact and energy savings in \cdc s, and presented the design and implementation of \shrink, a system that leverages this tradeoff to yield significant energy savings while affecting user-perceived response times in a controlled manner. Shrink's novel network-aware server consolidation algorithm reduces network energy use by up to 42\% compared to network-unaware server consolidation schemes. Our experiments based on content access traces from an Akamai datacenter showed that our energy optimization algorithms reduce energy use by 35\% compared to a conservative baseline scheme that provisions servers as per the peak demand while increasing mean response time by 8\% over it. Overall, our findings encourage deployment of energy optimization techniques  in \cdc s.

%We also studied the interaction between load balancing policies and server energy optimization in a \cdc\ and showed that a simple random load balancing policy is effective in avoiding load hotspots and results in a small impact on cache hit rates even as energy optimization algorithms vary the set of active servers in a \cdc.

%Further, we designed novel  algorithms that select the set of active servers and switches in a ``network-aware'' manner to provide additional network energy savings.

>>>>

%!TEX root = shrink.tex
\section{Conclusions}
Content datacenters used for storing and serving content to end-users are common today. A major barrier to  widespread adoption of server and network consolidation is \cdc\ operators' concern on the impact on SLAs or user-perceived response times. Our work takes a step towards addressing this concern by presenting a model to quantify the energy savings vs. response time inflation trade-off in \cdc s.  A key insight, supported via experiments, is that despite server consolidation, cache hit rates remain close to an unconsolidated datacenter,  which helps mitigate the impact of consolidation on user-perceived response times. We have designed and implemented \shrink, a system that leverages this tradeoff to yield significant energy savings while affecting user-perceived response times in a controlled manner. \shrink's novel network-aware server consolidation algorithm reduces network energy use by up to 42\% compared to network-unaware server consolidation schemes. \shrink, in experiments based on content access traces from an Akamai datacenter, reduced energy use by 35\% compared to a baseline scheme that keeps entire datacenter always on while increasing mean response time by 8\% over it. Overall, our findings encourage deployment of consolidation techniques to reduce \cdc\ energy use.

### 2015-07-17 shrink/intro.tex
Some edits to shrink chapter

Our consumption of digital content is increasing rapidly. The majority of the traffic on the Internet today is content, with video content by itself accounting for more than half the total traffic \cite{cisco-videogrowth, nielsen-video-growth}.  Not surprisingly, many datacenters  and  Points-of-Presence (PoPs) are dedicated to storing and serving content to end users. We call these datacenters and PoPs  {\em content datacenters} (or, \cdc s).

@@

%Energy use contributes to 15-20\% of total cost of ownership of a datacenter \cite{GreenbergCost, rasmussen2011determining, power-cost}.

%Yet, prior work on consolidation has given little attention to quantifying its impact on the end-user response time in a \cdc.  Work on server consolidation is either theoretical in nature with less emphasis on  empirical validation. Analytical work on server consolidation has measured the impact on response time based on theoretical utilization-vs.-response time models of servers with little empirical validation. Implementation-based work on server consolidation has focused largely on stateless servers whereas \cdc s maintain a large amount of state in the form of stored content;  the effect of consolidation on response time in \cdc s  cannot be evaluated accurately without accounting for the effect of consolidation on the availability of stored content. Work on network consolidation has focused mostly on evaluating traffic engineering metrics such as link utilization with little effort on evaluating response times for real applications.

%A key question for \cdc\ operators is the impact of consolidation on end-user response times.  
%
%An inflation in response times can cause SLA violations leading to revenue loss for operators.

%But, the end-user performance impact of consolidation for \cdc s has received little attention in prior work.  Analytical work on server consolidation, although based on theoretically sound principles, provides little empirical data on end-user response times. Implementation-based work on server consolidation has focused largely on stateless servers whereas \cdc s maintain a large amount of stored content as state;  the impact on response time in \cdc s  cannot be evaluated accurately without accounting for the impact on the availability of stored content. Work on network consolidation has focused mostly on measuring traffic engineering metrics such as link utilization with little effort on measuring response times for real applications.

The practicality of consolidation as an energy-saving tool for a \cdc\ operator depends on its impact on end-user response time. End-user response time is a key metric on which  operators are evaluated \cite{comparingCDN}. Further, a sharp response time inflation may be perceived as service unavailability, thereby causing a service-level agreement (SLA) violation and revenue loss for an operator \cite{amazonSLA, microsoftSLA, hpSLA}. Despite a large body of literature on consolidation schemes, quantifying the precise impact on user-perceived response times is not well understood today.
%To our knowledge, evaluating the impact of consolidation on the end-user response in a \cdc\ is still an open question.

@@

%Several factors can inflate response times
%
%Most prior on stateless servers, and a prior work on stateful datacenters storing content has considered very specific workload with small key-value items.

%As our work is grounded in implementation, it  accounts for several factors affecting response time - increased load on servers and network links, reduced storage and its impact on cache hit rates, non-steady state cache behavior due to on-off transitions, imperfect load balancing among servers - and hence, accurately quantifies the impact of consolidation on end-user response time.
%To provide the insights needed for an operator to reduce energy use via consolidation, we address these key questions:

%Our primary contribution in this paper is a model to relate the benefit of energy savings to the cost of response-time inflation for CDCs, and the design and implementation of Shrink, a system that aggressively leverages this relationship in order to yield significant savings in energy use in CDCs while affecting user-perceived response times in a controlled manner. 
Our primary contribution is to quantify the tradeoff between energy savings via consolidation and response times for a real \cdc's workload, and the design and implementation of a system, \shrink, to leverage this tradeoff.
\shrink\ reduces the energy use of servers and switches via consolidation, while enabling operators to achieve the desired response time and hardware reliability. \shrink\ uses a novel network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and increases the potential network energy savings. 
\shrink\ does not require changes to datacenter network fabric, depending only on ECMP (equal-cost multipath) for routing, support for which is widely available today.  As our work is grounded in implementation, it  accounts for several factors affecting response time - increased load on servers and network links, reduced storage and its impact on cache hit rates, non-steady state cache behavior due to on-off transitions, imperfect load balancing among servers - and hence, accurately quantifies the impact of consolidation on end-user response time.

>>>>

In today's content-dominated Internet \cite{cisco-videogrowth, nielsen-video-growth}, it is not surprising that many datacenters and Points-of-Presence (PoPs) are dedicated to storing and serving content to end users. We call these datacenters and PoPs {\em content datacenters} (or, \cdc s).

%Our consumption of digital content is increasing rapidly. The majority of the traffic on the Internet today is content, with video content by itself accounting for more than half the total traffic \cite{cisco-videogrowth, nielsen-video-growth}.  Not surprisingly, many datacenters  and  Points-of-Presence (PoPs) are dedicated to storing and serving content to end users. We call these datacenters and PoPs  {\em content datacenters} (or, \cdc s).

@@

The practicality of consolidation as an energy-saving tool for a \cdc\ operator depends on its impact on end-user response time. End-user response time is a key metric on which  operators are evaluated \cite{comparingCDN}. Further, a sharp response time inflation may be perceived as service unavailability, thereby causing a service-level agreement (SLA) violation and revenue loss for an operator \cite{amazonSLA, microsoftSLA, hpSLA}. Despite a large body of literature on consolidation schemes, quantifying the precise impact on user-perceived response times is not well understood today.

@@

Our primary contribution is to quantify the tradeoff between energy savings via consolidation and response times for a real \cdc's workload, and the design and implementation of a system, \shrink, to leverage this tradeoff. \shrink\ reduces the energy use of servers and switches via consolidation, while enabling operators to achieve the desired response time and hardware reliability. A novel aspect of \shrink\ is a network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and achieves greater network energy savings over network-unaware server consolidation schemes. Further, our server and network consolidation schemes require only ECMP (equal-cost multipath) support for routing, and hence, are deployable with existing datacenter network fabrics.

Our work is grounded in implementation, due to which it accounts for several factors affecting response time - increased load on servers and network links, reduced storage and its impact on cache hit rates, non-steady state cache behavior due to on-off transitions, imperfect load balancing among servers - and hence, accurately quantifies the impact of consolidation on end-user response time. A key insight, supported via experiments, is that cache hit rates, despite server consolidation, remain close to an unconsolidated datacenter. A small reduction in hit rates helps mitigate the impact of consolidation on response times. This finding is explained by the skewed content popularity obseved in real workloads, due to which working set size of content remains small compared to the storage available in a CDC.

%a simple placement strategy LRU caching achieves a cache hit rate near to an unconsolidated datacenter.

%A key insight we leverage is that due to skew in the content popularity  commonly observed is real content workloads cache hit rates are minimally affected despite consolidation. A simple LRU caching scheme achieves a cache hit rate near to an unconsolidated datacenter.
%, and substantiate via experiments,

### 2015-07-17 tepaper/section1.tex
Some edits to shrink chapter

While ISP traffic engineering schemes are commonly evaluated using link-utilization based metrics, these metrics are ill-suited to compare schemes based on user-perceived performance or their network capacity, i.e., the factor of surges in traffic demand that a network can tolerate, under location diversity. As discussed previously in Chapter \ref{ch:te-background}, link utilization-based metrics may not reflect application performance that depends on other metrics such as propagtion delay, access link capacity and backbone link capacity also. Further, link utilization based metrics, in particular MLU, do not reflect network capacity under location diversity as explained before.

@@

%Our approach to evaluating traffic engineering is empirical and application-centric.  Our experiments carefully and at scale simulate end-to-end application behavior. This approach enables us to evaluate TE schemes with respect to their impact on application performance in terms of TCP throughput for elastic traffic and a quality-of-service metric,  MOS score for VoIP quality \cite{MOS-formula}, for inelastic traffic. Further, this approach enables us to model application adaptation to location diversity and measure capacity using a novel capacity metric called Surge Protection Factor (SPF). SPF captures the factor of increase in demand that can be sustained while accounting for location diversity.

%Further, this approach enables us to simulate application adaptation to location diversity.

%Our research shows that link utilization based metrics, in particular the maximum link utilization (MLU) is a poor metric to evaluate user-perceived performance or \emph{network capacity}, i.e., the maximum surge in traffic demand that a network can support. Our results support our previous hypothesis that link utilization based metrics may not reflect user-perceived performance that depends on other factors also.
%Previously, we explained why link utilization based metrics may not reflect user-perceived performance that depends on other factors also.  Previously, we explained why MLU is no longer an effective metric of capacity in the presence of location diversity. In this Chapter, we propose a new capacity metric called Surge Protection Factor (SPF) for comparing traffic engineering schemes. SPF captures the factor of increase in demand that can be sustained while accounting for location diversity.

>>>>

While ISP traffic engineering schemes are commonly evaluated using link-utilization based metrics, these metrics are ill-suited to compare schemes based on user-perceived performance or network capacity, i.e., the factor of surge in traffic demand that a network can tolerate, under location diversity. As discussed previously in Chapter \ref{ch:te-background}, link utilization-based metrics may not reflect application performance that depends on other metrics such as propagtion delay, access link capacity and backbone link capacity also. Further, link utilization based-metrics, in particular MLU, do not reflect network capacity under location diversity as explained before.

@@
